---
title: "Coursera – Practical Machine Learning assignment"
author: "vleonov"
date: "21 Mar 2015"
output:
  html_document:
    keep_md: true
---

> ### Background
>
>Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 
>
>
> ###Data 
>
>The training data for this project are available here: 
>
>https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
>
>The test data are available here: 
>
>https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
>
>The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 
>
> ###What you should submit
>
>The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 
>
>1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
>2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 
>
> ###Reproducibility 
>
>Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis. 

#Intro

We will try to predict exercide class, using data recorded from several people using several sensors. We will user `Random Forest Modeling`, which gives us over 99% accuracy on the training set of 75% of the total data. Moreover, in `Random Forest` there is no need for cross-validation, because it estimated internally, see http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm#ooberr

#Prepare the enviroment

Load required libraries and set seed for reproducibility.

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}
library(data.table)
library(caret)
library(parallel)
library(doParallel)

set.seed(2103)
```

#Prepare the dataset

Read the training and testing (actually validating) data.

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}
data <- read.csv(file="data/pml-training.csv", head=TRUE, sep=",", na.strings = c("", "NA", "#DIV/0!"))
validation <- read.csv(file="data/pml-testing.csv", head=TRUE, sep=",", na.strings = c("", "NA", "#DIV/0!"))
```

We need only sensors' data for prediction. Moreover we exclude data, which contains NA values.

```{r}
isNA <- function(x) { any(is.na(x))}
hasNA <- sapply(data, excludeNA)

isPredictor <- !hasNA & grepl("belt|forearm|arm|dumbbell", names(hasNA))
predictors <- names(data)[isPredictor]
```
```{r, echo=FALSE}
print(predictors)
```

Subset the dataset to include only predictors and outcome variable `classe`.
```{r}
data <- data[,c('classe', predictors)]
```

Convert `classe` to factor.
```{r}
data$classe <- factor(data$classe)
```

It's time to divide dataset to training and testing partitions.
```{r}
inTrain <- createDataPartition(data$classe, p=0.75, list=FALSE)
training <- data[inTrain,]
testing <- data[-inTrain,]
```

Let's see what we have.
```{r}
summary(training)
```

Looks like we need to preprocess data. Do it by centering and scaling.
```{r}
preProc <- preProcess(training[,c(predictors)])
```

Apply this preprocessing to all datasets.
```{r}
training <- data.table(data.frame(classe = training$classe, predict(preProc, training[,predictors])))
testing <- data.table(data.frame(classe = testing$classe, predict(preProc, testing[,predictors])))
validation <- data.table(data.frame(problem_id = validation$problem_id, predict(preProc, validation[,predictors])))
```

#Training

It's a good idea to use all available cores and process random forests in parallel.
```{r, eval=FALSE}
cl <- makeCluster(detectCores() )
registerDoParallel(cl)
ctrl <- trainControl(classProbs=TRUE, savePredictions=TRUE, allowParallel=T)
```

Let's start computation. It will take several minutes, good time to drink a cup of cofee.
```{r, eval=FALSE}
modFit <- train(classe ~ ., method="rf", data=training, trControl=ctrl)
save(modFit, file="modFit.RData")
```
```{r, echo=FALSE}
load(file="modFit.RData")
```

Stop the clusters for parallel.
```{r, eval=FALSE}
stopCluster(cl)
```

#Testing

```{r, warning=FALSE, message=FALSE}
confusionMatrix(predict(modFit, testing), testing$classe)
```
Accuracy is 0.998 - very good.

#Validation

Write submission files to submit the result to Coursera.
```{r}
pml_write_files = function(x){
  n = length(x)
  path <- "predictionAssignment_files/answers"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predict(modFit, validation))
```

All of validation test are correct.